Evolution of \\ ice sheet geometry \\ using Stokes dynamics

### 2021-06-12 talk/talk.tex
minor edit
I could give this talk if necessary.

\documentclass[usepdftitle=false]{beamer}

@@

\titlegraphic{\includegraphics[width=0.7\textwidth]{figs/titleextruded.png} \\ \vspace{-3mm} \includegraphics[width=0.15\textwidth]{figs/uafbw.png}}

@@

\date{\phantom{foo} \bigskip \bigskip \bigskip \\ SIAM GS21 \\ with assistance from Lawrence Mitchell}

@@

@@

\begin{itemize}
\item[]
    \begin{itemize}
    \item $P_1$ because of low regularity at free boundary
    \item evaluate $\Phi^h(s^h)=-\bu|_{s^h}\cdot\bn_{s^h}$ using numerical solution of Stokes problem $F_{\Lambda_{s^h}}(\bu^h,p^h)=0$
    \end{itemize}
\item seeking a solver which is:
    \begin{itemize}
    \item \alert{robust}
    \item \alert{optimal} $O(m)$ work over $m$ degrees of freedom in $\mathcal{V}^h$
        \begin{itemize}
        \item[{\color{black} $\circ$}] requires multigrid
        \end{itemize}
    \item \emph{for now these are aspirations \dots}
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{columns}
\column{0.35\textwidth}
    \begin{itemize}
    \small
    \item which solution variable?
    \item choice between surface elevation ($s$) and thickness ($H$)
    \item interior smoothness means $s$ is better for multigrid
        \begin{itemize}
        \item need to \emph{not} have fake ice where $s=b$
        \end{itemize}
    \end{itemize}
\column{0.65\textwidth}
    \includegraphics[width=\textwidth]{figs/giscross.png}
\end{columns}
\end{frame}

\begin{frame}{multilevel constraint decomposition}

\begin{itemize}
\item FIXME
\end{itemize}

\includegraphics[width=0.5\textwidth]{figs/decompclassical.png}

\input{tikz/vcycle.tex}
\end{frame}

@@

\begin{itemize}
\item FIXME
\end{itemize}

\includegraphics[width=0.8\textwidth]{figs/extruded.png}
\end{frame}

\begin{frame}{smoothers}

\begin{itemize}
\item FIXME
\end{itemize}
\end{frame}

@@

\begin{frame}{Some references}
%\bibliography{../paper/msg.bib}
%\bibliographystyle{plain}

\begin{thebibliography}{1}
\bibitem{BuelerMitchell2022}
E.~Bueler and L.~Mitchell. \textbf{Multilevel computation of glacier geometry from Stokes dynamics.} in preparation

\hrulefill

\bibitem{Brinkerhoffetal2017}
D.~Brinkerhoff, M.~Truffer, and A.~Aschwanden. \textbf{Sediment transport drives tidewater glacier periodicity.} {\em Nature Communications}, 8(1):1--8, 2017.

@@

>>>>

\documentclass[usepdftitle=false,usenames,dvipsnames]{beamer}

@@

\titlegraphic{\includegraphics[width=0.7\textwidth]{figs/titleextruded.png} \\ \vspace{-8mm} \includegraphics[width=0.2\textwidth]{figs/uafbw.png}}

@@

\date{\phantom{foo} \bigskip \bigskip \bigskip \\ SIAM GS21}

@@

\begin{frame}{overview}
\Large
\begin{enumerate}
\item explain the problem I'm obsessed with
\item discuss the current barriers to success
\end{enumerate}

\normalsize
\vspace{10mm}
\begin{itemize}
\item<2> work in progress, with no results yet
    \begin{itemize}
    \item<2> low expectations, please
    \end{itemize}
\item<2> I am giving this talk at 4am my time
    \begin{itemize}
    \item<2> ditto
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item[]
    \begin{itemize}
    \item $P_1$ because of low regularity at free boundary
    \item evaluate $\Phi^h(s^h)=-\bu|_{s^h}\cdot\bn_{s^h}$ by solving Stokes problem $F_{\Lambda_{s^h}}(\bu^h,p^h)=0$
    \end{itemize}
\item seeking a solver which is:
    \begin{itemize}
    \item \alert{robust}
    \item \alert{near optimal} $O(m^{1+\eps})$ work over $m$ degrees of freedom in $\mathcal{V}^h$
        \begin{itemize}
        \item[{\color{black} $\circ$}] requires multigrid \dots more below
        \end{itemize}
    \item \emph{these are aspirations}
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{columns}
\column{0.35\textwidth}
    \begin{itemize}
    \small
    \item which solution variable?
    \item surface elevation ($s$) or thickness ($H$)?
    \item interior smoothness means $s$ is better for multigrid
    \item bed roughness means \emph{not} having ``fake ice'' where ice-free ($s=b$)
    \end{itemize}
\column{0.65\textwidth}
    \includegraphics[width=\textwidth]{figs/giscross.png}
\end{columns}
\end{frame}

\begin{frame}{multilevel constraint decomposition (MCD)}

\begin{block}{strategy}
decompose $\mathcal{K}^h$ using mesh hierarchy and then apply V-cycles
\end{block}

\bigskip
\begin{columns}
\column{0.52\textwidth}
\begin{itemize}
\item mesh levels over $\Omega$:

$\mathcal{T}^0 \text{ (coarse) to } \mathcal{T}^J$
\item given fine-level iterate $s^J$
\item \emph{defect constraint} $\chi^J = b^J - s^J$
\item \emph{monotone restriction} $R^\oplus$

\item defect constraints on levels:

$\chi^j = R^\oplus \chi^{j+1}$
\item admissible perturbations:

$z^j \ge \chi^j \iff s^j + z^j \ge b^j$
\end{itemize}
\column{0.48\textwidth}
\includegraphics[width=1.05\textwidth]{figs/decompclassical.png}
\end{columns}
\end{frame}

\begin{frame}{multilevel constraint decomposition (MCD)}

\begin{block}{strategy}
decompose $\mathcal{K}^h$ using mesh hierarchy and then apply V-cycles
\end{block}

\begin{itemize}
\item this strategy gives $O(m\log m)$ solver for Laplacian obstacle problem (Tai, 2003)
    \begin{itemize}
    \item using Gauss-Seidel smoother
    \item \only<1>{a multilevel VI strategy is the only hope for (near) optimality}\only<2>{{\color{ForestGreen} a multilevel VI strategy is the only hope for (near) optimality}}
    \item for nonlinear VI, need additional FAS strategy {\Large\strut} \only<2>{{\color{ForestGreen} \, {\Large $\checkmark$}}}
    \end{itemize}
\item for IGP a single residual evaluation requires solving Stokes
    \begin{itemize}
    \item expensive but implemented {\Large\strut} \only<2>{{\color{ForestGreen} \, {\Large $\checkmark$}}}
    \item Stokes is nonlocal

\smallskip
    \item Jacobian dense

\smallskip
    \item pointwise smoothers (GS, Jacobi) use diagonal entries
        \begin{itemize}
        \item<2>[{\color{black} $\circ$}] \alert{which apparently fail {\Large \, :-(}}
        \end{itemize}
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{center}
\includegraphics[width=0.8\textwidth]{figs/extruded.png}
\end{center}

\begin{itemize}
\item for each base mesh level $\mathcal{T}^j$ in FAS V-cycle,
\item evaluate the residual $\Phi(s^j) - a^j$ in the VI:
    \begin{itemize}
    \item source $a^j$ arises from FAS formula
    \item extrude to current geometry $s^j$ (with or without cliffs)
    \item no extrusion where ice free\footnote{thanks to Lawrence Mitchell for enabling this Firedrake capability}
    \item solve Stokes problem $F_{\Lambda_{s^j}}(\bu^j,p^j)=0$ (Isaac et al, 2015)
    \item compute $\Phi(s^j) = - \bu^j|_{s^j} \cdot \bu_{s^j}$ and subtract $a^j$
    \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}{smoother options}

\begin{itemize}
\item \only<1>{now all the evil is in the smoother}\only<2>{\alert{now all the evil is in the smoother}}
\item options from weaker to stronger:
    \begin{itemize}
    \item nonlinear Richardson?
        \begin{itemize}
        \item[{\color{black} $\circ$}] not promising because degenerate
        \end{itemize}
    \item nonlinear pointwise smoothers (GS, Jacobi)?
        \begin{itemize}
        \item[{\color{black} $\circ$}] \only<1>{extremely expensive because residual nonlocal} \only<2>{\alert{extremely expensive because residual nonlocal}}
        \item[{\color{black} $\circ$}] seem to fail anyway
        \end{itemize}
    \item nonlinear \texttt{PCPatch} smoothers?
        \begin{itemize}
        \item[{\color{black} $\circ$}] what does this even mean?
        \end{itemize}
    \item exploit SIA analogy?
        \begin{itemize}
        \item[{\color{black} $\circ$}] pointwise smoothers work for SIA
        \end{itemize}
    \item reduced-space Newton using banded Jacobian approximation?
        \begin{itemize}
        \item[{\color{black} $\circ$}] finite-difference Jacobian entries using pseudo-coloring
        \end{itemize}
    \end{itemize}

\bigskip
\item<2> status report: \alert{nothing works well so far}
\end{itemize}
\end{frame}

\begin{frame}{conclusion}

\begin{itemize}
\item \emph{GOAL 1}:\, no shallow approximations in flow physics {\Large \strut} \only<2-3>{{\Large \color{ForestGreen} $\checkmark$}}
    \begin{itemize}
    \item<2-3> my Stokes solver is working in Firedrake on extruded mesh
    \item<2-3> solver choices by (Isaac et al, 2015) putatively optimal
    \end{itemize}
\item \emph{GOAL 2}:\, beat explicit time-stepping and enable paleo {\Large \strut} \only<3>{\Large \alert{:-(}}
    \begin{itemize}
    \item<2-3> problem is
        \begin{itemize}
        \item<2-3>[{\color{black} $\circ$}] constrained (nonlinear VI on $\mathcal{K} = \{s\ge b\}$)
        \item<2-3>[{\color{black} $\circ$}] diffusive in the large (Stokes $\approx$ SIA)
        \item<2-3>[{\color{black} $\circ$}] nonlocal (Stokes)
        \end{itemize}
    \item<2-3> {\color{ForestGreen} MCD FAS path to optimality for the VI}
    \item<3> \alert{stuck on finding a robust optimal smoother}
    \end{itemize}

\normalsize
\bigskip\bigskip
\item<3> \emph{thanks for your attention \dots questions?}
\end{itemize}
\end{frame}

@@

\begin{frame}{Some references 1}
%\nocite{*}
%\bibliography{../paper/msg.bib}
%\bibliographystyle{plain}

\begin{thebibliography}{1}
\bibitem{Brinkerhoffetal2017}
D.~Brinkerhoff, M.~Truffer, and A.~Aschwanden. \textbf{Sediment transport drives tidewater glacier periodicity.} {\em Nature Communications}, 8(1):1--8, 2017.

\bibitem{Bueler2016}
E.~Bueler. \textbf{Stable finite volume element schemes for the shallow-ice approximation.} {\em J. Glaciol.}, 62 (232):230--242, 2016.

@@

\bibitem{JouvetBueler2012}
G.~Jouvet and E.~Bueler. \textbf{Steady, shallow ice sheets as obstacle problems: well-posedness and finite element approximation.} {\em SIAM J. Appl. Math.}, 72(4):1292--1314, 2012.

\bibitem{JouvetRappaz2011}
G.~Jouvet and J.~Rappaz. \textbf{Analysis and finite element approximation of a nonlinear stationary {S}tokes problem arising in glaciology.} {\em Advances in Numerical Analysis}, 2011.
\end{thebibliography}
\end{frame}

\begin{frame}{Some references 2}
\begin{thebibliography}{1}
\bibitem{Tai2003}
X.-C. Tai. \textbf{Rate of convergence for some constraint decomposition methods for nonlinear variational inequalities.} {\em Numer. Math.}, 93(4):755--786, 2003.
